Cheminformatics: The Computer Science of Chemical Discovery

### 2011-07-24 paper.tex
Added in Jannas edits

\maketitle
\begin{abstract}
  One of the most prominent success stories in all of science over the last decade has
  been the advance of bioinformatics: interdisciplinary collaboration between
  computer scientists and molecular biologists led to the sequencing the human genome
  (among many other accomplishments), because researchers connected biological
  concepts to the theory of string algorithms. However, despite this great
  success, few computer scientists are familiar with a related (and older!)
  discipline---cheminformatics, the use of computers to represent
  small molecules structures and analyze their properties based on aspects
  of their structures. Cheminformatics has a wide applicability
  from the discovery of novel drug candidates to optimizing
  certain physicochemical properties of small molecules. Until
  recently, the data and techniques of chemoinformatics have been closely guarded
  secrets of companies whose financial success depended on being the first to produce
  the new ``miracle molecule.'' Only within the last decade---and largely because of
  chemists volunteering their time for an Open Science ``movement''---do researchers
  now have access to freely available software packages and databases of tens of
  millions of chemicals. As a result, chemists now confront a variety of unsolved
  algorithmic problems that could not have been tackled a decade ago, but whose
  solutions are critical to research ranging from determining the behavior of small
  molecules in biological pathways, to finding a cure for malaria.
\end{abstract}

\category{H.4}{Information Systems Applications}{Miscellaneous}
\terms{
\begin{itemize}
\item \textit{Tautomerism} - Tautomers are isomers (same molecular formula but different structural formula) of organic compounds that readily interconvert by a chemical reaction called tautomerization.
\item \textit{Protonation and protonation state} - In chemistry, protonation is the addition of a proton (H$^+$) to an atom, molecule, or ion. The different states (different number of H$^+$ or on different atoms) are called protonation states for a molecular compound (e.g. de-protonated, mono-protonated, di-protonated, or any variations of those states).
\item \textit{Bioassay} - A system measuring the biological activity of a chemical compound on a particular biological assay (with all its parameters).
\item \textit{Metabolites} - Metabolites are the intermediates and products of metabolism. The term metabolite is usually restricted to small molecules (chemical compounds).
\item \textit{Mass spectrometry} - Mass spectrometry (MS) is an analytical technique that measures the mass-to-charge ratio of charged particles. It is used for determining masses of particles, for determining the elemental composition of a sample or molecule, and for elucidating the chemical structures of molecules, such as peptides and other chemical compounds.
\item \textit{Ligand (coordination chemistry)} - an atom, ion, or functional group that donates one or more of its electrons through a coordinate covalent bond to one or more central atoms or ions.
\item \textit{Ligand (biochemistry)} - a substance that binds to a protein.
\item \textit{Receptor (biochemistry)} - in biochemistry, a protein molecule that receives and responds to a neurotransmitter, or other substance.
\item \textit{Proteochemometric modeling} - combining data mining for chemical compounds (ligands) and proteins within the same machine learning framework.
\item \textit{Biodegradation} - is the chemical dissolution of materials by bacteria or other biological means.
\item \textit{Reaction catalysis} - Catalysis is the change in rate of a chemical reaction due to the participation of a substance called a catalyst.
\item \textit{Conformation and conformational isomerism} - In chemistry, conformational isomerism is a form of stereoisomerism in which the isomers can be interconverted exclusively by rotations about formally single bonds. Single possible states of compounds are called conformers.
\item \textit{SMARTS} - SMiles ARbitrary Target Specification (SMARTS) is a language for specifying substructure patterns in molecules
\end{itemize}
}
\keywords{cheminformatics, graphs, chemistry, databases, knowledge management, open source, open data}

\section{Is Cheminformatics the New Bioinformatics?}

Novel technologies in the life sciences churn out information at an ever
increasing rate---public data stores such as at the European Bioinformatics
Institute (EBI) contain on the order of 10 petabytes of biological information.
However, while biological information has been available publicly for decades,
the same has not been true for chemical information until very recently. It was
not until 2004 that a large public small molecule structure repository (PubChem)
became freely available. This was followed by other databases, as discuss later
in this article. In the same vein, open source software and the publication of
algorithms have only recently become a main focus of the field
\cite{faulon2010}.

@@

Making chemical information public is important because the drugs we
take are the result of intensive and costly research, and the more
information is shared (a process sometimes very hard to achieve in
pharmaceutical companies) the easier it should be to develop novel
treatments. How chemical information relates to computer science
becomes clear when realizing the amount of data available
currently---74 million different molecules are currently stored in
various public databases (such as PubChem) with a large amount of
annotations, and the design of data structures for the representation
and manipulation of chemical structure, as well as subsequent data
mining of this information, are areas where expertise in computer
science is dearly needed. (For a recent, detailed introduction to
cheminformatics, written for computer scientists, see
\cite{brown2009}).

Cheminformatics comprises different areas, which broadly speaking can be divided
into three fields: \emph{capturing data} (using lab notebooks or potentially using
formats such as the Chemical Markup Language for publications); \emph{storing data}
(designing database schemas, devising ontologies) and \emph{mining data} (such as for
bioactivity prediction, which might involve algorithms based on graphs).
However, a major difference from the bioinformatics field is the type of
information being analyzed: While bioinformatics often deals with sequences, the
domain of cheminformatics is chemical structures. In case of the former,
information can very frequently be represented as one-dimensional strings, which
are relatively easy to handle computationally. In the case of the latter, chemical
structures may possess rings, branches as well as multiple valid representations
of even the same structure, potentially giving rise to ambiguities. Hence,
chemical structure is often thought to be more difficult to standardize.

To illustrate the difficulty of standardizing chemical structures, different
representations of the same chemical structure are shown in Figure 1. The top
left hand corner shows a structure a chemist would draw it---every `corner'
representing a carbon atom (which can also be represented as a C), every O
denoting oxygen, every N nitrogen, every H hydrogen, and so on. This is the
standard format chemists use when exchanging information with their peers.
However, this format is not directly amenable to data mining---different atoms might
be collapsed into one node, sets of atoms might have the same relevant
properties assigned. Hence, the top right hand corner of Figure 1 shows an
annotated graph representation, taking into account that atoms, though being of
different elemental atom type, can have similar properties---\emph{e.g.}, both oxygen
and nitrogen carry an `On' label in this graph.

@@

Graph mining algorithms can be applied to this type of annotated graph
(employing large databases of thousands of molecules), to discover
patterns in large-scale datasets.  Many successful applications of
this concept have been published \cite{wegner2006,horst2009}. Still,
this ``annotated graph'' representation is far from being the end of
the story. Molecules are three-dimensional entities, as represented in
the bottom left hand corner of the figure, and the precise 3D shape is
relevant in many situations (although not in others).  Nevertheless,
the 3D information is lost when one only considers the graph of a
molecule. To extend the concept of three-dimensionality further, in
many biological systems the \emph{surface properties} of the molecule
are what appear to be relevant, as shown in the bottom right hand
corner.  Hence, while graph mining on chemicals is certainly a
promising avenue to explore, one needs to keep in mind that not all
molecular properties are encoded in the graph and thus often other
`descriptors' need to be used to capture the essentials of a chemical
system \cite{bender2004}.

From the above it becomes clear that the nature of many problems in
cheminformatics require that a complex mixture of combinatorial and
continuous techniques.  If we oversimplify the nature of chemical
structures and their properties, we can not describe the world around
us in sufficient detail. On the other hand, if we try to encode every
detail (\emph{e.g.}, all possible 3D structures and their properties,
all metabolites of a structure in humans in different tissues), the
complexity of collecting, storing, and mining data increases
dramatically. For the field of cheminformatics to flourish, we need
closer collaboration between chemists (or, more generally, life
scientists) and computer scientists---where the former need to be able
to pose their problems in a way relevant for practical applications;
and where the latter are able to devise ways of capturing, storing,
and analyzing chemical data by bridging the gap between risk in
generalizable encodings and complexity (space and time). The goals of
cheminformatics are to support better chemical decision making by
1. storing and integrating data in maintainable ways, 2.  providing
enough open standards and tools so software engineering concepts allow
applications and data to be used across heterogeneous platforms, and
3. mining the many chemical property spaces in a time- and
space-efficient way. We will discuss all three goals in the following
sections.

\subsection{Outline}
\label{sec:outline}

The remainder of the article discusses various topics in
cheminformatics by placing them in the context of the relevant
subfields of computer science, aiming to highlight the challenges that
need to be addressed and possible entry points for computer science
research. Section \ref{sec:databases} discusses issues related to the
storage of chemical structures in relational databases and how
ontologies allow us to query such collections beyond the standard
graph based subgraph matching approaches. While databases usually
contain explicit molecules, enumeration of sctructures based on
various constraints is key to a variety of chemical problems ranging
from mass spectrometry to reaction planning. Section
\ref{sec:struct-enum} discusses challenges in this area. Section
\ref{sec:activity-mining} highlights the problem of making predictions
of biological activity from chemical structure and tieing it all
together, Section \ref{sec:knowledge-management} discusses the
challenges of managing chemical structure data along with associated
biological data in production scenarios. Finally Section
\ref{sec:role-open-source} discusses the public sources of chemical
information and tools available for the handling of chemical structre
-- both of which are required for for the development of
cheminformatics research programs.

@@

Chemical graphs can differ based on the manner in which the structures
are drawn. Also chemical behaviors can result in multiple encodings of
``the same'' molecule and its tautomeric states, different protonation
states, or many variations of implicit chemical expert
knowledge. Checking whether two graphs represent the same chemical is
an expensive operation (the next section will discuss this in-depth);
thus, a representation of the structure that is unique and invariant
with respect to how the structure is drawn is needed for
identification. An early effort in this direction was the Simplified
Molecular Input Line Specification (SMILES), and specifically the
generation of a single canonical SMILES for a particular structure. However,
with time several
different SMILES implementations became available with no universal
standard of generating a canonical SMILES. After years of research and various
commercial products, there is today no single public standard for a
line-notation of molecular labeled graphs, which is available to a
broad audience (academic and
commercial). %The very same problem does also exist for
%bioassay ontologies and the lack of options to normalize the property spaces of
%chemical compounds, e.g. biological assays or clinical data outcome
%(biobanking).

With each database using their own algorithm for molecular encoding,
the automated exchange of data between different databases was not
facilitated. As more and more data became available online,
structure-based identification became a pressing need, and the IUPAC
International Chemical Identifier (InChI) code was developed. The
InChI is a non-proprietary, structured textual identifier for chemical
entities [1]. InChI identifiers are not intended to be read and
understood by humans, but are useful for computational matching of
chemical entities. For rapid database lookups, the InChIKey is a
hashed key for the InChI, which has an invariant length of 14
characters. Figure 1 illustrates the SMILES, InChI and InChIKey for
lipoic acid.

%Figure 1: The standard chemical graph illustration together with the SMILES,
%InChI and InChIKey codes for lipoic acid. Note that the SMILES is much more
%human-readable than the others.

The InChI is now widely used for matching identical chemical structures, but it
is still limited. For example, it cannot differentiate between
certain types of stereoisomers: Figure 2 illustrates two stereoisomers for which
the generated InChI is the same.

%Figure 2: The generated InChI is the same for the stereoisomers cisplatin and
%transplatin.

@@

Enumerating molecules is not only an interesting academic exercise but has
practical utility as well. The foremost application of enumeration is \emph{structure
elucidation} of natural compounds such as metabolites. 
Ideally, the wishful bench chemist collects experimental data
from an unknown compound, the data is fed into a code, and the resulting
structure is returned. Although that streamlined process may not yield a unique
solution, there are commercial software products such as MOLGEN
(http://molgen.de) that can, for instance, list all structures matching a given
molecular formula, or mass spectrometry (MS) data. Another important application
of enumeration is \emph{molecular design}. Here the problem is to design compounds
(drugs, for instance) that optimize some physical, chemical, or biological
property or activity. Although less prolific and more recent than structure
elucidation, molecular design has introduced some novel solutions to molecular
enumeration. Finally, with the advent of combinatorial chemistry, molecular
enumeration has taken a central role as it allows computational chemists to
construct virtual libraries, test hypotheses, and provides guidance in the
design of optimal combinatorial experiments.

@@

The underlying computational complexity and potential intractability
of molecular enumeration may not be a bottleneck after all, as
computational chemists have developed and successfully used
enumeration tools to generate large chemical libraries like GDB-13
comprising almost a trillion chemical
structures\footnote{http://www.dcb-server.unibe.ch/groups/reymond/gdb/home.html}. Yet,
the current enumeration software products do not generally produce
stereoisomers or tautomers; these require specific enumeration
procedures, which are still the subject of investigations by the
cheminformatics community.

@@

The first step in the predictive modeling of biological activities is
to compare compounds. Either we transform molecular graphs into a
numerical vector representation (a.k.a. \emph{descriptors} or
\emph{features}) of chemical structures, or we use algorithms for
comparing molecular graphs directly (\emph{molecule kernel}). In the
first case thousands of descriptors (many of which are equivalent)
have been described \cite{todeschini2000},
and a variety of tools are available for their calculation. Given a
set of molecules, observed activities and a descriptor vector, the
problem can be considered a traditional statistical modeling
problem. In the second case no explicit vector representation is
required, which prevents the molecular graph to vector transformation
challenge (loss-less encoding risk). On the other hand imposes the
implicit kernel encoding that we have to compare all compounds against
all compounds, which imposes large-scale mining challenge (time and
space complexity risk). Note that the public PubChem dataset alone
contains over 30 million chemical compounds and over 70 million other
substances.

@@

With the advent of high-throughput screening technologies, large
libraries of compounds can be screened against multiple targets in an
efficient manner. Such panel assays provide a broad, systems-level
view of small molecule activities.  Models developed on such data
afford us the opportunity to identify targets, characterize off-target
effects and so on. However, most approaches to this problem are rather
simplistic since they develop multiple individual models.  Instead one
could imagine an approach that takes into account the correlations in
observed activities against multiple targets within a single
model. Such an approach should lead to more robust predictions for
panel assays. Finally, does this also better reflect clinical relevant
compound profiles \cite{kuhn2010} and a 'personalized medicine'
concept (e.g. drug-drug interaction profiles).

@@

The relevance of scientific knowledge management is increasing not only for
reducing the risk in current research, but also for enabling new
collaboration/innovation opportunities with internal/external and fast growing
number of external/public partners. In fact, in cheminformatics and chemistry,
scientists already switched many years ago from paper lab-notebooks to
online collaboration platforms, called ELNs (Electronic Lab Notebooks). So, why
have chemists and companies already adopted to ``Enterprise 2.0,'' a social
online collaboration culture? Many other disciplines are still busy with solving
ROI (return-of-investment), risk reduction, opportunity creation, change
culture, and technical collaboration challenges for scientists and scientific
data.

@@

ELNs are typically much more accepted and used within the cheminformatics
domain than in the bioinformatics domain, which is picking up speed lately. The
question is why? In chemistry every experiment requires a lot of expert
knowledge of chemists. Therefore, approaches for reducing risks by communicating
with fellow chemists about similar reactions or compounds is critical. This
requires capturing the reaction details, compound details, and time consuming
analysis ensuring quality data, which can be searched by others.

It is worth mentioning that chemistry is traditionally a field where many people
have been used to huge bookshelves of reactions and compounds. Chemists changed
not only to online ELN collaboration for getting more efficient, but also
because it allowed collaboration with trusted colleagues. So, third party paper
catalogs and academic publications proved to be
inefficient, and even more critically, they did not take compounds of trusted
colleagues into account. Overall the change of management to ELNs, similar to
``Enterprise 2.0,'' required to deliver on the promise to increase
interconnectivity, serve as silo-breaker, and to improve collaboration
opportunities, and it performed these functions many years ago, already.

The future still leaves many challenges ahead. Using external/public databases
with chemical and bioactivity data remains a challenge due to differences in
ontologies, synchronization and maintenance, and efficient
substructure and similarity search within complex data types
(compounds). A
collaboration with external parties, e.g. contract research, poses other
problems with duplicate checking for compounds (e.g. tautomeric or salt
structures), efficient data synchronization, most often a double-registration is
required for fulfilling the data quality processes on both sides. If external
partners are small they often do not even have sufficient IT resources
themselves and rely on external services themselves. Cloud services could help
not only to provide a service infrastructure for all parties involved,
but also to provide the required private/public access toolbox. Furthermore
there are still many data sources not being indexed properly, since chemistry
patents are often cryptic and image/text mining remains a puzzle. One could
argue that the image/text mining in scientific journals is less challenging; in
fact, it remains a huge challenge and many chemists do not trust automated
non-expert validated mining sources. The closed CAS (Chemical Abstract Service)
is a highly trusted source, and public chemistry and bioactivity space has to
improve quality and interconnectivity. Efforts like ChEMBL and PubChem-BioAssay
are on the right track. Still, improving data quality and standards between
public and closed sources will be absolutely critical for ensuring a constant
growth, usage, and collaboration scenarios between private and public parties.
Only when the technical and scientific questions are solved can we ensure that
legal questions about contribution metrics will get enough room to grow.

@@

It is also important to note that certain challenges faced by nearly
all cheminformatics toolkit stem from the graph representation of
chemistry. This representation, while succinct and amenable to
computation, is only an approximation of reality---and edges cases
abound. Existing areas of interest are enumeration of colored graphs,
taking into account symmetry, and symmetry detection itself, which not
merely derives from the chemical graph, but typically also from the 3D
geometry of the chemical. However, more challenging is that chemical
representation of a molecule really needs to capture non-static
graphs, in order to take into account delocalisation and tautomeric
phenomena molecules can practically undergo. It is important to
realize that though chemical representations have increased in
sophistication over the years, many chiral molecules cannot be
satisfactorily handled by current representation systems.

Given that there is ample scope for software engineering (algorithm
implementation, code quality and analysis, build systems etc), both
the CDK and Open Babel are open to new contributions. Both are driven
via public, active mailing lists and public issue tracking systems.
Development questions are actively and openly discussed to arrive at a
consensus and contributors are encouraged to get involved with any of
the following areas: development of new features, improvement of
existing features, performance optimization, fixing bugs,
documentation, website development, reporting bugs, and helping users
on the mailing list. Anyone interested should email the respective
developer's mailing list (either cdk-devel@lists.sf.net or
openbabel-devel@lists.sf.net) with their interests and background, and
a mentor will be assigned to help them join the project.

>>>>

\maketitle
\begin{abstract}
b  One of the most prominent success stories in all the sciences over the last decade has
  been the advance of bioinformatics: interdisciplinary collaboration between
 computer scientists and molecular biologists led to the sequencing of the human genome
  and many other accomplishments, since researchers connected biological
  concepts to the theory of string algorithms. However, despite this great
  success, few computer scientists are familiar with a related (and older!)
  discipline---cheminformatics, the use of computers to represent
 the structures of small molecules and analyze their properties. 
  Cheminformatics has a wide applicability,
  from the discovery of novel drug candidates to the optimization of
  certain physicochemical properties of small molecules. Until
 recently, the data and techniques of cheminformatics have been closely guarded
  secrets of companies whose financial success depended on being the first to produce
  the new ``miracle molecule.'' Only within the last decade---and largely because of
 chemists volunteering their time for an Open Science ``movement''---have researchers
  gained access to freely available software packages and databases of tens of
  millions of chemicals. Academic chemists now confront a variety of unsolved
  algorithmic problems that could not have been tackled a decade ago, but whose
  solutions are critical to research ranging from determining the behavior of small
  molecules in biological pathways, to finding a cure for malaria.
\end{abstract}

\category{H.4}{Information Systems Applications}{Miscellaneous}
\terms{
\begin{itemize}
\item \textit{Tautomerism} - Tautomers are isomers (same molecular formula but different structural formula) of organic compounds that readily interconvert by a chemical reaction called tautomerization.
\item \textit{Protonation and protonation state} - In chemistry, protonation is the addition of a proton (H$^+$) to an atom, molecule, or ion. The different states (different number of H$^+$ or on different atoms) are called protonation states for a molecular compound (e.g. de-protonated, mono-protonated, di-protonated, or any variations of those states).
\item \textit{Bioassay} - A system measuring the biological activity of a chemical compound on a particular biological assay (with all its parameters).
\item \textit{Metabolites} - Metabolites are the intermediates and products of metabolism. The term metabolite is usually restricted to small molecules (chemical compounds).
\item \textit{Mass spectrometry} - Mass spectrometry (MS) is an analytical technique that measures the mass-to-charge ratio of charged particles. It is used for determining masses of particles, for determining the elemental composition of a sample or molecule, and for elucidating the chemical structures of molecules, such as peptides and other chemical compounds.
\item \textit{Ligand (coordination chemistry)} - an atom, ion, or functional group that donates one or more of its electrons through a coordinate covalent bond to one or more central atoms or ions.
\item \textit{Ligand (biochemistry)} - a substance that binds to a protein.
\item \textit{Receptor (biochemistry)} - in biochemistry, a protein molecule that receives and responds to a neurotransmitter, or other substance.
\item \textit{Proteochemometric modeling} - combining data mining for chemical compounds (ligands) and proteins within the same machine learning framework.
\item \textit{Biodegradation} - is the chemical dissolution of materials by bacteria or other biological means.
\item \textit{Reaction catalysis} - Catalysis is the change in rate of a chemical reaction due to the participation of a substance called a catalyst.
\item \textit{Conformation and conformational isomerism} - In chemistry, conformational isomerism is a form of stereoisomerism in which the isomers can be interconverted exclusively by rotations about formally single bonds. Single possible states of compounds are called conformers.
\item \textit{SMARTS} - SMiles ARbitrary Target Specification \\(SMARTS) is a language for specifying substructure patterns in molecules
\end{itemize}
}
\keywords{cheminformatics, graphs, chemistry, databases, knowledge management, open source, open data}

\section{Is Cheminformatics the New \\Bioinformatics?}

Novel technologies in the life sciences churn out information at an ever
increasing rate---public data stores such as at the European Bioinformatics
Institute (EBI) contain in the order of 10 petabytes of biological information.
However, while biological information has been publicly available for decades,
the same has not been true for chemical information until very recently. It was
not until 2004 that a large public small molecule structure repository (PubChem)
became freely available. This was followed by other databases, as discussed later
in this article. In the same vein, open source software and the publication of
algorithms have only recently become a main focus of the field
\cite{faulon2010}.

@@

Making chemical information public is important because the
drugs we take are the result of intensive and costly research, 
which is facilitated by the extent of information available. The
more information is shared (a process very hard to achieve
in pharmaceutical companies), the more information is available to each 
researcher, facilitating the development of
novel treatments. The relevance of computer
science in this research becomes clear in considering the quantity of data available
currently---74 million different molecules are currently stored in
public databases (such as PubChem), together with an even larger number of annotations.
The design of data structures for this chemical information, as well as subsequent
data mining of this information, are areas where expertise in computer
science is urgently needed. (For a recent, detailed introduction to
cheminformatics, written for computer scientists, see
\cite{brown2009}).

Cheminformatics comprises different areas, which broadly speaking can be divided
into three fields: \emph{capturing data} (using lab notebooks or potentially using
formats such as the Chemical Markup Language for publications); \emph{storing data}
(designing database schemas, devising ontologies) and \emph{mining data} (such as for
bioactivity prediction, which might involve algorithms based on graphs).
However, a major difference from bioinformatics lies in the type of
information being analyzed: While bioinformatics often deals with sequences, the
domain of cheminformatics is chemical structures. In the case of the former,
information can very frequently be represented as one-dimensional strings, which
are relatively easy to handle computationally. In the case of the latter, chemical
structures may possess rings, branches as well as multiple valid representations
of the same structure, potentially giving rise to ambiguities. Hence,
chemical structure is often reported to be more difficult to standardize.

To illustrate the difficulty of standardizing chemical structures, different
representations of the same chemical structure are shown in Figure 1. The top
left hand corner shows a structure a chemist would draw it---every `corner'
representing a carbon atom (which can also be represented as a C), every O
denoting oxygen, every N nitrogen, every H hydrogen, and so on. This is the
standard format chemists use when exchanging information with their peers.
However, this format is not directly amenable to data mining---different atoms might
be collapsed into one node, and sets of atoms might have the same relevant
properties assigned. Hence, the top right hand corner of Figure 1 shows an
annotated graph representation, taking into account that atoms, though being of
different elemental atom type, can have similar properties---\emph{e.g.}, both oxygen
and nitrogen carry an `On' label in this graph.

@@

Graph mining algorithms can be applied to this type of annotated graph
(employing large databases of thousands of molecules), to discover
patterns in large-scale datasets.  Many successful applications of
this concept have been published \cite{wegner2006,horst2009}. Still,
this ``annotated graph'' representation is far from being the end of
the story. Molecules are three-dimensional entities, as represented in
the bottom left hand corner of the figure, and the precise 3D shape is
relevant in many situations (although not in others).  Nevertheless, the
3D information is lost when one only considers the graph of a
molecule. To extend the concept of three-dimensionality further, in
many biological systems the \emph{surface properties} of the molecule
are what appear to be relevant, as shown in the bottom right hand
corner.  Hence, while graph mining on chemicals is certainly a
promising avenue to explore, one needs to keep in mind that not all
molecular properties are encoded in the graph (for example, neither
rotational states, nor different charge states are encoded) and thus
often other `descriptors' need to be used to capture the essentials of
a chemical system \cite{bender2004}.

From the above it becomes clear that the nature of chemical problems
is often a complex mixture of combinatorial and continuous
problems. If we oversimplify the nature of chemical structures and
their properties, we can not describe the world around us in
sufficient detail. On the other hand, if we try to encode every detail
(\emph{e.g.}, all possible 3D structures and their properties, all
metabolites of a structure in humans in different tissues), the
complexity of collecting, storing, and mining these data increases
dramatically. For the field of cheminformatics to flourish, we need
closer collaboration between chemists (or, more generally, life
scientists) and computer scientists---where the former need to be able
to pose their problems in a way relevant for practical applications;
and where the latter are able to devise ways of capturing, storing,
and analyzing chemical data which achieve optimal balances between the different risks in 
generalizable encodings and complexity (space and time). The goals of cheminformatics are to
support better chemical decision making by 1. storing and integrating
data in maintainable ways, 2.  providing enough open standards and
tools so software engineering concepts allow applications and data to
be used across heterogeneous platforms, and 3. mining the many
chemical property spaces in a time- and space-efficient way. We will
discuss all three goals in the following sections.

@@

Chemical graphs can differ based on the manner in which the structures
are drawn. Also chemical behaviors can result in multiple encodings of
``the same'' molecule and its tautomeric states, different protonation
states, or many variations of implicit chemical expert
knowledge. Checking whether two graphs represent the same chemical is
an expensive operation (the next section will discuss this in-depth);
thus, a representation of the structure that is unique and invariant
with respect to how the structure is drawn is needed for
identification. An early effort in this direction was the Simplified
Molecular Input Line Specification (SMILES), and specifically the
generation of a single canonical SMILES for a particular structure. However,
with time several
different SMILES implementations became available with no universal
standard of generating a canonical SMILES. 
%After years of research and various
%commercial products, there is today no single public standard for a
%line-notation of molecular labeled graphs, which is available to a
%broad audience (academic and
%commercial). %The very same problem does also exist for
%bioassay ontologies and the lack of options to normalize the property spaces of
%chemical compounds, e.g. biological assays or clinical data outcome
%(biobanking).

With each database using their own algorithm for molecular encoding,
automated exchange of data between different databases was hindered. 
As more and more data became available online,
structure-based identification became a pressing need, and the IUPAC
International Chemical Identifier (InChI) code was developed. The
InChI is a non-proprietary, structured textual identifier for chemical
entities \cite{inchi}. InChI identifiers are not intended to be read and
understood by humans, but are useful for computational matching of
chemical entities. For rapid database lookups, the InChIKey is a
hashed key for the InChI, which has an invariant length of 14
characters. Figure \ref{smiles} illustrates the SMILES, InChI and InChIKey for
lipoic acid.

%Figure {smiles}: The standard chemical graph illustration together with the SMILES,
%InChI and InChIKey codes for lipoic acid. Note that the SMILES is much more
%human-readable than the others.

The InChI is now widely used for matching identical chemical structures, but it
is still limited. For example, it cannot differentiate between
certain types of stereoisomers: Figure \ref{cistrans} illustrates two stereoisomers for which
the generated InChI is the same.

%Figure {cistrans}: The generated InChI is the same for the stereoisomers cisplatin and
%transplatin.

@@

Enumerating molecules is not only an interesting academic exercise but has
practical utility as well. The foremost application of enumeration is \emph{structure
elucidation} of natural compounds such as metabolites. 
Ideally, the wishful bench chemist collects experimental data
from an unknown compound, the data is fed into a code, and the resulting
structure is returned. Although that streamlined process may not yield a unique
solution, there are commercial software products such as MOLGEN
(http://molgen.de) that can, for instance, list all structures matching a given
molecular formula, or mass spectrometry (MS) data. Another important application
of enumeration is \emph{molecular design}. Here the problem is to design compounds
(drugs, for instance) that optimize some physical, chemical, or biological
property or activity. Although less prolific and more recent than structure
elucidation, molecular design has introduced some novel solutions to molecular
enumeration. Finally, with the advent of combinatorial chemistry, molecular
enumeration has taken a central role as it allows computational chemists to
construct virtual libraries and test hypotheses, and it provides guidance in the
design of optimal combinatorial experiments.

@@

The underlying computational complexity and potential intractability
of molecular enumeration may not be a bottleneck after all, as
computational chemists have developed and successfully used
enumeration tools to generate large chemical libraries such as GDB-13,
comprising almost a trillion chemical
structures\footnote{http://www.dcb-server.unibe.ch/groups/reymond/gdb/home.html}. Yet,
the current enumeration software products do not generally produce
stereoisomers or tautomers; these require specific enumeration
procedures, which are still the subject of investigations by the
cheminformatics community.

@@

The first step in the predictive modeling of biological activities is
to compare compounds. Either we transform molecular graphs into a
numerical vector representation (a.k.a. \emph{descriptors} or
\emph{features}) of chemical structures, or we use algorithms for
comparing molecular graphs directly (\emph{molecule kernel}). In the
first case thousands of descriptors (many of which are equivalent)
have been described \cite{todeschini2000},
and a variety of tools are available for their calculation. Given a
set of molecules, observed activities and a descriptor vector, the
problem can be considered a traditional statistical modeling
problem. In the second case no explicit vector representation is
required, which prevents the challenge of transforming the molecular 
graph into a vector without losing information. On the other hand, the 
implicit kernel encoding imposes the challenge that we have to compare all compounds against
all compounds, which imposes large-scale mining challenge, which risks overrunning available
processing time and space. Note that the public PubChem dataset alone
contains over 30 million chemical compounds and over 70 million other
substances.

@@

With the advent of high-throughput screening technologies, large libraries of
compounds can be screened against multiple targets in an efficient manner. Such
panel assays provide a broad, systems-level view of small molecule activities.
Models developed on such data afford us the opportunity to identify targets,
characterize off-target effects and so on. However, most approaches to this
problem are rather simplistic since they develop multiple individual models.
Instead one could imagine an approach that takes into account the correlations
in observed activities against multiple targets within a single model. Such an
approach should lead to more robust predictions for panel assays. Finally, 
this also better reflects clinically relevant compound profiles \cite{kuhn2010}
and a `personalized medicine' concept (e.g. drug-drug interaction profiles) [The
Decision Tree].

@@

The relevance of scientific knowledge management is increasing not only for
reducing the risk in current research, but also for enabling new
collaboration/innovation opportunities with internal partners and a rapidly growing
number of external/public partners. In fact, in cheminformatics and chemistry,
scientists already switched many years ago from paper lab-notebooks to
online collaboration platforms, called ELNs (Electronic Lab Notebooks). So, why
have chemists and companies already adopted ``Enterprise 2.0,'' a social
online collaboration culture? Many other disciplines are still busy with solving
ROI (return-of-investment), risk reduction, opportunity creation, change
culture, and technical collaboration challenges for scientists and scientific
data.

@@

ELNs are typically much more accepted and used within the cheminformatics
domain than in the bioinformatics domain, which is picking up speed lately. The
question is why? In chemistry, every experiment requires a lot of expert
knowledge. Therefore, approaches for reducing risks by communicating
with fellow chemists about similar reactions or compounds is critical. This
requires capturing the reaction details, compound details, and time consuming
analysis details which ensure quality data, such that these data can be searched by others.

It is worth mentioning that chemistry is traditionally a field where many people
have been used to huge bookshelves of reactions and compounds. Chemists changed
to online ELN collaboration not only to increase efficiency, but also
because it allowed collaboration with trusted colleagues. Third party paper
catalogs and academic publications proved to be
inefficient, and even more critically, they did not take compounds of trusted
colleagues into account. Overall the change of management to ELNs, similar to
``Enterprise 2.0,'' required delivering on the promise to increase
interconnectivity, serve as silo-breaker, and to improve collaboration
opportunities, and indeed it already performed these functions many years ago.

However, the future still leaves many challenges ahead. Using external/public databases
with chemical and bioactivity data remains a challenge due to differences in
ontologies, synchronization and maintenance, and efficient
substructure and similarity search within complex data types
(compounds). A
collaboration with external parties, e.g. contract research, poses other
problems with duplicate checking for compounds (e.g. tautomeric or salt
structures) and efficient data synchronization, and most often a double-registration is
required for fulfilling the data quality processes on both sides. If external
partners are small they often do not even have sufficient IT resources
themselves and thus rely on external services. Cloud services could help
not only to provide a service infrastructure for all parties involved,
but also to provide the required private/public access toolbox. Furthermore,
there are still many data sources not being indexed properly, since chemistry
patents are often cryptic and image/text mining remains a puzzle. One could
argue that the image/text mining in scientific journals is less challenging; in
fact, it remains a huge challenge and many chemists do not trust automated
non-expert validated mining sources. The closed CAS (Chemical Abstract Service)
is a highly trusted source, and public chemistry and bioactivity space has to
improve quality and interconnectivity to compare with it. Efforts like ChEMBL and PubChem-BioAssay
are on the right track. Still, improving data quality and standards between
public and closed sources will be absolutely critical for ensuring constant
growth, usage, and collaboration scenarios between private and public parties.
Only when the technical and scientific questions are solved can we ensure that
legal questions about contribution metrics will get enough room to grow.

@@

It is also important to note that certain challenges faced by nearly
all cheminformatics toolkits stem from the graph representation of
chemistry. This representation, while succinct and amenable to
computation, is only an approximation of reality---and edge cases
abound. Existing areas of interest are enumeration of colored graphs,
taking into account symmetry, and symmetry detection itself, which not
merely derives from the chemical graph, but typically also from the 3D
geometry of the chemical. However, more challenging is that chemical
representation of a molecule really needs to capture non-static
graphs, in order to take into account delocalisation and the tautomeric
phenomena molecules can undergo in practice. It is important to
realize that although chemical representations have increased in
sophistication over the years, many chiral molecules cannot be
satisfactorily handled by current representation systems.

Given that there is ample scope for software engineering (algorithm
implementation, code quality and analysis, build systems etc), both
the CDK and Open Babel are open to new contributions. Both are driven
via public, active mailing lists and public issue tracking systems.
Development questions are actively and openly discussed to arrive at a
consensus and contributors are encouraged to get involved with any of
the following areas: development of new features, improvement of
existing features, performance optimization, fixing bugs,
documentation, website development, reporting bugs, and helping users
on the mailing list. Anyone interested should email the respective
developers' mailing list (either cdk-devel@lists.sf.net or
openbabel-devel@lists.sf.net) with their interests and background, and
a mentor will be assigned to help them join the project.
